Direct observation of Feshbach enhanced \swave scattering of fermions

### 2015-04-21 draft1.tex
Started on 2nd round of Ian's comments
Got most of the way through the intro

\title[]{Direct observation of \swave scattering of fermions crossing a Feshbach resonance}

@@

\ead{spielman@umd.edu}

\begin{abstract}
We directly mapped \swave scattering across a Feshbach resonance in\K. In this experiment, we collided a pair of degenerate clouds and directly imaged the scattered atoms. Owing to their low density, few atoms scattered, even near the resonance. To optimize signal to noise, we develop techniques to interpret absorption imaging in the regime where the optical intensity changes dramatically as light traverses the cloud, and recoil induced 
detuning corrections are significant. We apply these techniques to our \swave scattering data to extract an estimate for the resonant magnetic
field value. These techniques are generally applicable, and can be used to observe effective p-wave scattering in the presence of spin-orbit coupling in a spin polarized  Fermi gas.

@@

\vspace{2pc}
\noindent{\it Keywords}: Quantum gases, Atomic physics, Something else

@@

\section{Introduction}
Feshbach resonances provide a stand-by technique for tuning the interaction strength in ultracold atomic gases. Even weak interactions play a crucial role in the physics of atomic Bose-Einstein Condensates (BECs), giving rise to, for example, their characteristic Thomas-Fermi density profiles. The effects of interactions in Fermi gases, however, are harder to observe. The density of Fermi clouds is reduced by a factor of $~10^3$ from that of BECs, making it necessary to enhance the strength of interactions in order to observe them. Feshbach resonances alter the inter-atomic scattering length via the magnetic field. Not only do they enable detection of interactions, but they make it possible to tune interactions from attractive to repulsive, allowing for the phase transition from the BCS to BEC regime at sufficiently low temperatures \cite{RegalThesis}. 
\par A Feshbach resonance occurs when a bound molecular state of two atoms energetically approaches the free atomic state \cite{Chin10}. The energy of the atomic states is defined by their hyperfine state and tunable with a bias magnetic field. The Feshbach resonance can thus be approached by changing the bias field. The effect of the resonance on the scattering length between two free atoms in the appropriate hyperfine state is given by 
\begin{equation}
a(B)=a_{bg}\left(1-\frac{\Delta}{B-B_0}\right),
\label{feshbachEq}
\end{equation}
where $a$ is the scattering length, $a_{bg}$ is the scattering length away from the resonance, $\Delta$ is the width of the resonance, and $B_0$ is the field value at which the resonance occurs. Note that the scattering length tends to infinity from either side of the resonance.
\par To use Feshbach resonances as a tool it  is necessary to have a good measurement of the parameters in the above equation.  The exact value of the resonant field value $B_0$ is impossible to calculate analytically and must be estimated via numerical models or determined experimentally. Some experimental techniques that have been used include observation of loss due to three-body inelastic scattering, measurements of re-thermalization timescales, or measurements of anisotropic expansion, all of which infer the elastic scattering cross section from collective behavior of the cloud \cite{Regal03,OHara02,Monroe93}. 
\par Here we present a new technique for measuring the location and width of a Feshbach resonance. We collide two clouds of a degenerate Fermi gas and directly image the resulting \swave scattering halo as a function magnetic field strength. This allows us to directly observe the enhancement in scattering without relying on collective behavior. We measured the fraction of atoms scattered during the collision, and from this fraction deduced the resonant magnetic field  and width of the resonance.
%\par The techniques developed in this experiment for observing Fermion scattering can be extended to engineering higher order partial wave interactions, as has been done for bosons \cite{Williams2012}.
\par Our Fermi gases are so dilute that even with Feshbach resonance enhancement of the scattering cross section, very few atoms will scatter as a fraction of total atom number. This makes direct detection of \swave scattering halos difficult due to detection uncertainty, which disproportionately effects low atom numbers. In order to optimize the signal to noise for low atom numbers, we entered an imaging regime that relies on simulation for an accurate interpretation of the images. We performed these simulations and used the results to extract the atom number and the scattered fraction from our images.
\par This paper is in two parts. In the first, we study absorption imaging in the presence of significant Doppler shift and show how we use our results to interpret data. In the second, we describe our\swave scattering experiment and extract a measure of the location of the Feshbach resonance in \K.

@@

\par To obtain an absorption image, one shines an on or near resonant probe beam ($\delta\approx0$) onto the atomic cloud. Some of the light is scattered by the atoms, and the part of the beam that made it through the cloud is imaged onto a camera, as seen in Figure \ref{fig:absorptionIntor}a (top). Then, the atoms are allowed to leave the trap, and the probe light is shined directly at the camera to calibrate the intensity of light that the atoms saw (bottom). The intensity in the second camera image is called the initial intensity, $I_{0}$, as it is assumed to be the same as the intensity in the first beam before interaction with atoms. The intensity in the first image is called the final intensity, $I_f$.  

@@

\par The observed intensity in the two images can be used to infer the number of atoms that the light encountered. To do this, consider what happens to the light as it travels along the imaging axis $z$ through a column of atoms. In general, the atomic density is three dimensional, $\rho(x,y,z)$. For the purposes of this discussion, let us focus on a single pixel of the camera, and thus a single value of $I_0$ and $I_f$ and a single column of atoms, $\rho(z)$. We will not be able to infer the entire atomic distribution, but for each pixel we will be able to deduce the total 2-d atomic density, $n = \int \rho\left(z\right) \,\mathrm{d}z$. As the light travels through a column of atoms, each atom will scatter light according to eq. \ref{eq:scatrate}. Therefore, the atoms further along the imaging axis $z$ will see a reduced light intensity, as seen in Figure \ref{fig:absorptionIntor}b. The amount of light lost to scattering as a function of $z$ is given by

@@

In the limit where the probe intensity is much smaller than the saturation intensity, $I_0\ll I_{sat}$, this becomes simply $-\ln \left(I_f/I_0\right)$. This quantity is called the optical depth, or OD. Since this is the simplest possible model that relates the observed intensities to the column density $n$, we call this quantity $OD_0$. Once the probe intensity becomes comparable to the saturation intensity, the second term in eq.[\ref{eq2}] becomes significant. We can then call this the optical density corrected for high probe intensity, or

@@

\par When an atom absorbs a photon from the laser light field and gets excited to a higher energy level, by conservation of momentum it must also acquire a velocity in the direction of the light field. This is called the recoil velocity, and it is given by $v_r=\hbar k/m$, where $k$ is the wavenumber of the light field and $m$ is the atomic mass. When the photon is subsequently emitted and the atom returns to its ground state, it is emitted in an arbitrary direction. Thus, the momentum lost due to emitting a photon is on average zero. The variance, however, is not zero, allowing the atoms to acquire some momentum transverse to the laser field. We will ignore this correction, but the effect of this on the atomic cloud is picture in Figure \ref{fig:absorptionIntor}c. 
\par Once the atoms absorb enough photons, they acquire a substantial velocity in the direction of the light field, and this velocity Doppler shift them from the light field itself. Thus, an atom that absorbs $N$ photons acquires a velocity of $N v_r$ and a detuning of $\delta=k N v_r$ from the probe beam. Therefore, even if the probe beam is initially on resonance with the atomic transition, we cannot neglect the detuning term in the scattering rate as time goes on. Furthermore, this detuning will vary both with imaging time $t$ and with distance along the propagation direction $z$ (Figure \ref{fig:detunedBlobs}). Thus, the intensity lost to the atoms also acquires a time dependence: 

@@

\par One can consider this equation perturbatively in time and obtain corrections to second order in imaging time, $OD_2$  \cite{LJLthesis}. However, the perturbative treatment breaks down shortly after the zeroth order approximation of eq. [\ref{eq2}] (Figure \ref{fig:ODcorrections}). In order to adequately correct for the recoil induced detuning of the atoms, we must simulate the process and obtain numerical predictions for $I_f$ given a certain imaging time, atomic density, and probe intensity. 

@@

\begin{algorithm}
\caption{Stationary atom model}
\label{algorithm1}
\begin{algorithmic}
\STATE $I[n=0,t]=I_0$ \COMMENT{$n$ is the bin index, $t$ is the time index, $I$ is in units of $I_{sat}$} 
\STATE $\delta[n, t=0]=0$ \COMMENT{light initially resonant, $\delta$ in units of $\Gamma/2$}
\STATE $I_f=0$
\FOR[loop over time steps]{$t=0$ to $t_f$}
 \FOR[loop over bins, N is total bin number]{$n=1$ to $N$}
 \STATE $A=\sigma_0\rho[n] dz$
 \STATE $B=v_r dt/(\hbar c \rho[n])$ 
\STATE $I[n,t]=I[n-1,t] - A I[n-1,t]/(1+\delta[n,t-1]^2+I[n-1,t])$  \COMMENT{Eq.\ref{eq3}}
\STATE $\delta[n,t]=\delta[n,t-1]+B\left(I[n-1,t]-I[n,t]\right)$  \COMMENT{Eq.\ref{eq4}} 
\ENDFOR 
\STATE $I_f  \mathrel{+}= I[N,t]dt$
\ENDFOR
\STATE $OD_{corr1}=-\ln{(I_f/I_0t_f)}$
\end{algorithmic}
\end{algorithm}

\par To check the validity of our simulation, we can take it to certain limits in which the problem becomes analytically solvable. In the limit that the probe intensity is much weaker than saturation, $I_0\ll I_{sat}$, the atoms will not absorb enough light to significantly detune from resonance. We can then neglect time dependence and eq.[\ref{eq3}] reduces to 
\begin{equation}
\frac{dI(z)}{dz}=-\rho\sigma_0 I(z),
\end{equation}
from which we recover the simple
\begin{equation}
\sigma_0 n = OD_0 = -\ln I_0/I_f. \label{eq6}
\end{equation}
In the limit that the probe intensity is much stronger than saturation, $I_0\gg I_sat$, even far detuned atoms will absorb light at their maximum, allowing us to again neglect the time dependence and reduce \ref{eq3} to 
\begin{equation}
\frac{dI(z)}{dz}=-\rho\sigma_0 I_{sat}, 
\end{equation}
which integrates to 
\begin{equation}
\sigma_0 n = \frac{I_0 - I_f}{I_{sat}}. \label{eq8}
\end{equation}
We recognize the right hand sides of eq.[\ref{eq6}] and eq.[\ref{eq8}] as the two terms in the expression for $OD_1$, eq.[\ref{eq2}]. Thus, in both limits $OD_{corr1}$ should coincide with the value of $OD_0$ obtained by fixing an input atom number $n$ in eq.[\ref{eq2}]. Indeed, they coincide as seen in Figure \ref{fig:IsatLimits}.
\begin{figure}
	\includegraphics{figure3.pdf}
\caption{Optical density as a function of probe intensity as predicted by the simulation (blue dots) and by eq.[\ref{eq2}] (green lines), for three different imaging times. The predictions agree in both the high and low intensity limits, and differ for probe intensities comparable to the saturation intensity. The difference is enhanced with increased imaging time.}  
\label{fig:IsatLimits}
\end{figure}
\par We can use the results of this simulation to check if the stationary atom assumption is self consistent, ie if the distance traveled by the atoms in one bin during the imaging time is less than the bin size. However, as can be seen from Figure \ref{fig:atomTravel}, it turns out that not only do the atoms travel more than the bin size, but they also travel more than the size of the whole cloud, and the back of the cloud even overtakes the front for long enough imaging times. Thus, the atomic distribution as a function of position changes dramatically during the imaging time, and our stationary assumption is completely invalid. 
\begin{figure}
	\includegraphics{figure4.pdf}
\caption{Position of atoms as a function of imaging time for atoms in the first, middle, and last bins of the simulation. The probe intensity here is $1.2 I_{sat}$, and the optical depth is 1.6.}  
\label{fig:atomTravel}
\end{figure}

@@

\STATE $I[n,t]=I[n-1,t] - A I[n-1,t]/(1+S[n].\delta^2+I[n-1,t])$  \COMMENT{Eq.\ref{eq3}}
\STATE $S[n].\delta\mathrel{+}=B\left(I[n-1,t]-I[n,t]\right)$  \COMMENT{Eq.\ref{eq4}, detuning in units of $\Gamma/2$} 

@@

\par First, we check the velocity behavior in this model against known limits. One such limit is that of a single supeatom. In this case, there is no attenuation, and the intensity seen by the superatom is constant at $I_0$. The only thing that evolves in time is the detuning of the single superatom. We can differentiate both sides of eq. [\ref{eq4}] to obtain

@@

and eq.[\ref{eq3}] simplifies to

@@

Combining eq.[\ref{eq9}] and eq.[\ref{oneAtom}], and converting to a dimensionless form, we obtain

@@

where $\Delta = 2\delta/\Gamma$, $\tilde{I} = I_0/I_{sat}$ and $k$ is the photonic wavevector. Eq.[\ref{eq10}] can be solved numerically, and is in good agreement with our simulation, as seen in Figure \ref{fig:oneAtomVel}.

@@

\par We use this model to study the time evolution of the cloud shape during imaging. This can be visualized as the phase space evolution of superatoms, as seen in Figure \ref{fig:phaseSpace}. We see that the cloud shape is actually strongly distorted during the imaging time. 

@@

\par We compare the optical depths predicted by each of the two models, $OD_{corr1}$ and $OD_{corr2}$. We find that, depite the significant changes in atomic distribution during the imaging time, the predicted optical densities are so slightly changed by including this effect that it is practically undetectable, as seen in Figure \ref{fig:atomTravel}. In fact, the difference $\left|OD_{corr1}-OD_{corr2}\right|/OD_{corr1} \le 0.005$. Thus, for the purposes of deducing the atom density from experimental optical depths, simply using a stationary model is sufficient. Furthermore, the actual atomic distribution $\rho(z)$ is largely irrelevant, and the only observable is the 2-D atomic density $n=\int\rho(z)\mathrm{d}z$. 

@@

\subsection{Signal to noise optimization}
This simulation can be used to interpret experimentally obtained initial and final intensities. For a given imaging time, we create a look-up table of predicted optical depth as a function of probe intensity and atomic density. We can then find the observed optical depth on this table, with the given probe intensity, and infer the atomic density. The uncertainty in the measured intensities can be propagated through this procedure, and we can establish optimal imaging parameters to maximize the signal to noise ratio of this detection scheme. 
\par Here, the only source of measurement uncertainty we consider is the photon shot noise, which is Poisson distributed and has a standard deviation proportional to $\sqrt{N_p}$, where $N_p$ is the photon number. We then propagate this uncertainty through our correction scheme to get the uncertainty in our deduced value of $\sigma_0 n$. We define the signal to noise ratio (SNR) as $\sigma_0 n/\delta_{\sigma_0 n}$, where $ \delta_{\sigma_0 n}$ is the propagated measurement uncertainty.
\par As seen in figure \ref{fig:SNR}, after about 40\us{} extending the imaging time no longer yields appreciable improvement in signal to noise ratio. The advantage of imaging for 40\us{} as opposed to 10\us{} where the uncorrected model is appropriate is about a factor of  1.5 increase in SNR. We performed the experiments described in the second section at 40\us{} imaging time.
\begin{figure}
	\subfigure[]{\includegraphics{figure9a.pdf}}
	\subfigure[]{\includegraphics{figure9b.pdf}}
\caption{Signal to noise ratio (SNR) for three different optical depths after correcting for recoil induced detuning. a. SNR as a function of probe intensity $I_0/I_{sat}=5.0$ and b. SNR as a function of probe intensity for an imaging time of 50\us{}.}  
\label{fig:SNR}
\end{figure}

\subsection{Calibration of saturation intensity}
A key component of good detection is a well calibrated measurement apparatus. In this case, the measurement apparatus is a CCD camera used to take the absorption images. We used Point Grey's Flea3 camera.  Each camera pixel  converts tho photons it is exposed to, with some efficiency, into photoelectrons, and then to a digital signal, and integer we will call the number of 'counts'. The number of photons that hit a pixel and the number of 'counts' it outputs are directly proportional. However, the proportionality constant depends on many factors, such as the quantum efficiency of the camera, the photoelectric conversion factor of the camera, and the polarization of the probe light. 
\par The most accurate way to determine this factor is through a direct experiment. In the limit where the system is adequately described by eq.[\ref{eq6}], only the ratio of the initial and final intensities matters, and thus this proportionality constant is irrelevant. In all other regimes, however, the ratio of the initial and final intensities to the saturation intensity also comes into play, making the proportionality constant significant. One way to approach this calibration is to determine the value of the saturation intensity in 'counts' per unit time. 
\par In order to calibrate the saturation intensity in camera counts per unit time, we take absorption images of a cloud of $^{40}K$ atoms at three different imaging times, 40us, 100us, and 200us, at varying probe intensities. We pick a small square in the center of the cloud, where the atomic density is approximately uniform, and average the initial and final intensities of each pixel in the square. Thus, for each image we obtain one $I_0$ and one $I_f$, in counts per microsecond. We then do a least squares fit of $OD_{corr}$, our simulated optical depth, to the data. The two fit parameters are the atomic density $n$ at the center of the cloud and the value of $I_{sat}$ in counts per microsecond. As seen in Figure \ref{fig:isatCalib}, the model produces a good fit to the experimental data, and we obtain a calibration of the saturation intensity for our experiment. 
\begin{figure}
	\includegraphics{figure8.pdf}
\caption{The optical depth as a function of probe intensity for three imaging times: $t=40$\us{} (blue),  $t=75$\us{} (green),  $t=100$\us{} (red). The dots represent experimental data and the lines represent the best fit of simulated data. The optimal fit parameters pictured are a $\sigma_0 n$ of 1.62 and saturation intensity of 29 counts/\us{}. }  
\label{fig:isatCalib}
\end{figure}

@@

\par We first corrected each image for recoil induced detuning using the simulation described above. We simulated a look-up table of predicted optical depths as a function of atomic column density and probe intensity. From our two absorption images we knew the probe intensity and optical depth for each pixel. We then used the simulated look-up table in an inverted fashion to deduce the atomic column density. An example of the effect of this correction procedure in shown in Figure \ref{fig:SampleCorrection}.  To improve the signal and compensate for our shot to shot number fluctuations, we took 15 equivalent images for each data point and, after correcting, added them together.

>>>>

\title[]{Direct observation of Feshbach enhanced \swave scattering of fermions}

@@

\ead{ian.spielman@nist.gov}

\begin{abstract}
We directly measured the \swave{} scattering cross-section of \K atoms accross a Feshbach resonance. We collided a pair of degenerate clouds and imaged the scattered atoms. Owing to their low density, few atoms scattered, even near the resonance. To optimize signal to noise, we develop techniques to interpret absorption imaging in the regime where the optical intensity changes dramatically as light traverses the cloud, and recoil induced 
detuning corrections are significant. We applied these techniques to our \swave{} scattering data and extracted the resonant magnetic
field value. These imaging techniques are generally applicable, and can be used to observe effective \pwave{} scattering in the presence of spin-orbit coupling in a spin polarized  Fermi gas.

@@

\vspace{2pc}
\noindent{\it Keywords}: Quantum gases, Atomic physics

@@

\section{Introduction}
Feshbach resonances are a stand-by technique for tuning the interaction strength in ultracold atomic gases. Even weak interactions play a crucial role in the physics of atomic Bose-Einstein Condensates (BECs), giving rise to, for example, their characteristic Thomas-Fermi density profiles. The effects of interactions in Fermi gases, however, are harder to observe. The density of Fermi clouds is typically reduced by a factor of $~10^3$ from that of BECs, making it necessary to enhance the strength of interactions in order to observe them. Feshbach resonances alter the inter-atomic scattering length via the magnetic field. Not only do they enable detection of interactions, but they make it possible to tune interactions from attractive to repulsive, allowing for the phase transition from the BCS to BEC regime at sufficiently low temperatures \cite{RegalThesis}. 
\par A Feshbach resonance occurs when a bound molecular state of few atoms energetically approaches the free atomic state \cite{Chin10}. The energy of the atomic states is defined by their hyperfine state and tunable with a bias magnetic field. The Feshbach resonance can thus be approached by changing the bias field. The effect of the resonance on the scattering length between two free atoms in the appropriate hyperfine states is
\begin{equation}
a(B)=a_{bg}\left(1-\frac{\Delta}{B-B_0}\right),
\label{feshbachEq}
\end{equation}
where $a$ is the scattering length, $a_{bg}$ is the scattering length away from the resonance, $\Delta$ is the width of the resonance, and $B_0$ is the field value at which the resonance occurs. Note that the scattering length tends to infinity from either side of the resonance.
\par To use Feshbach resonances as a tool it  is necessary to have a good measurement of the parameters in the above equation.  The exact value of the resonant field value $B_0$ is impossible to calculate analytically and must be estimated via numerical models \cite{Lysebo09, Gao11} or determined experimentally. Some experimental techniques that have been used to characterize Feshbach resonances include: the observation of loss due to three-body inelastic scattering, re-thermalization timescales, or anisotropic expansion, all of which infer the elastic scattering cross section from collective behavior of the cloud \cite{Regal03,OHara02,Monroe93}. 
\par Here we present a new technique for measuring the location and width of a Feshbach resonance. We collide a pair of ultra-cold Fermi gases and directly image the resulting \swave scattering halo as a function magnetic field strength. This allows us to observe the enhancement in scattering without relying on proxy effects. We measured the fraction of atoms scattered during the collision, and from this fraction deduced the resonant magnetic field  and width of the resonance.
%\par The techniques developed in this experiment for observing Fermion scattering can be extended to engineering higher order partial wave interactions, as has been done for bosons \cite{Williams2012}.
\par Our Fermi gases are so dilute that even with the resonant enhancement of the scattering cross section, only a small fraction of the atoms scatter. This makes direct detection of \swave scattering halos difficult due to detection uncertainty, which disproportionately affects regions of low atomic density. To optimize the signal to noise for low atom numbers, we utilized a non standard imaging regime. In this regime, simulation was necessary for an accurate interpretation of the images. We performed these simulations and used the results to extract the atom number and the scattered fraction from our images.
\par This paper is in two parts. In the first, we study absorption imaging in the presence of significant time-dependent Doppler shift and show how we use our results to interpret data. In the second, we describe our\swave scattering experiment and extract a measure of the location of the Feshbach resonance in \K.

@@

\par To obtain an absorption image, one shines an on or near resonant probe beam ($\delta\approx0$) onto the atomic cloud. Some of the light is scattered by the atoms, and the part of the beam that made it through the cloud is imaged onto a camera, as seen in Fig. \ref{fig:absorptionIntor}a (top). Then, the atoms are allowed to leave the trap, and the probe light is shined directly at the camera to calibrate the intensity of light that the atoms saw (bottom). The intensity in the second camera image is called the initial intensity, $I_{0}$, as it is assumed to be the same as the intensity in the first beam before interaction with atoms. The intensity in the first image is called the final intensity, $I_f$.  

@@

\par The observed intensity in the two images can be used to infer the number of atoms that the light encountered. To do this, consider what happens to the light as it travels along the imaging axis $z$ through a column of atoms. In general, the atomic density is three dimensional, $\rho(x,y,z)$. For the purposes of this discussion, let us focus on a single pixel of the camera, and thus a single value of $I_0$ and $I_f$ and a single column of atoms, $\rho(z)$. We will not be able to infer the entire atomic distribution, but for each pixel we will be able to deduce the total 2-d atomic density, $n = \int \rho\left(z\right) \,\mathrm{d}z$. As the light travels through a column of atoms, each atom will scatter light according to Eq. (\ref{eq:scatrate}). Therefore, the atoms further along the imaging axis $z$ will see a reduced light intensity, as seen in Fig. \ref{fig:absorptionIntor}b. The amount of light lost to scattering as a function of $z$ is given by

@@

In the limit where the probe intensity is much smaller than the saturation intensity, $I_0\ll I_{sat}$, this becomes simply $-\ln \left(I_f/I_0\right)$. This quantity is called the optical depth, or OD. Since this is the simplest possible model that relates the observed intensities to the column density $n$, we call this quantity $OD_0$. Once the probe intensity becomes comparable to the saturation intensity, the second term in Eq. (\ref{eq2}) becomes significant. We can then call this the optical density corrected for high probe intensity, or

@@

\par When an atom absorbs a photon from the laser light field and gets excited to a higher energy level, by conservation of momentum it must also acquire a velocity in the direction of the light field. This is called the recoil velocity, and it is given by $v_r=\hbar k/m$, where $k$ is the wavenumber of the light field and $m$ is the atomic mass. When the photon is subsequently emitted and the atom returns to its ground state, it is emitted in an arbitrary direction. Thus, the momentum lost due to emitting a photon is on average zero. The variance, however, is not zero, allowing the atoms to acquire some momentum transverse to the laser field. We will ignore this correction, but the effect of this on the atomic cloud is picture in Fig. \ref{fig:absorptionIntor}c. 
\par Once the atoms absorb enough photons, they acquire a substantial velocity in the direction of the light field, and this velocity Doppler shift them from the light field itself. Thus, an atom that absorbs $N$ photons acquires a velocity of $N v_r$ and a detuning of $\delta=k N v_r$ from the probe beam. Therefore, even if the probe beam is initially on resonance with the atomic transition, we cannot neglect the detuning term in the scattering rate as time goes on. Furthermore, this detuning will vary both with imaging time $t$ and with distance along the propagation direction $z$ (Fig. \ref{fig:detunedBlobs}). Thus, the intensity lost to the atoms also acquires a time dependence: 

@@

\par One can consider this equation perturbatively in time and obtain corrections to second order in imaging time, $OD_2$  \cite{LJLthesis}. However, the perturbative treatment breaks down shortly after the zeroth order approximation of Eq. (\ref{eq2}) (Fig. \ref{fig:ODcorrections}). In order to adequately correct for the recoil induced detuning of the atoms, we must simulate the process and obtain numerical predictions for $I_f$ given a certain imaging time, atomic density, and probe intensity. 

@@

\begin{algorithm}
\caption{Stationary atom model}
\label{algorithm1}
\begin{algorithmic}
\STATE $I[n=0,t]=I_0$ \COMMENT{$n$ is the bin index, $t$ is the time index, $I$ is in units of $I_{sat}$} 
\STATE $\delta[n, t=0]=0$ \COMMENT{light initially resonant, $\delta$ in units of $\Gamma/2$}
\STATE $I_f=0$
\FOR[loop over time steps]{$t=0$ to $t_f$}
 \FOR[loop over bins, N is total bin number]{$n=1$ to $N$}
 \STATE $A=\sigma_0\rho[n] dz$
 \STATE $B=v_r dt/(\hbar c \rho[n])$ 
\STATE $I[n,t]=I[n-1,t] - A I[n-1,t]/(1+\delta[n,t-1]^2+I[n-1,t])$  \COMMENT{Eq. (\ref{eq3})}
\STATE $\delta[n,t]=\delta[n,t-1]+B\left(I[n-1,t]-I[n,t]\right)$  \COMMENT{Eq. (\ref{eq4})} 
\ENDFOR 
\STATE $I_f  \mathrel{+}= I[N,t]dt$
\ENDFOR
\STATE $OD_{corr1}=-\ln{(I_f/I_0t_f)}$
\end{algorithmic}
\end{algorithm}

\par To check the validity of our simulation, we can take it to certain limits in which the problem becomes analytically solvable. In the limit that the probe intensity is much weaker than saturation, $I_0\ll I_{sat}$, the atoms will not absorb enough light to significantly detune from resonance. We can then neglect time dependence and Eq.(\ref{eq3}) reduces to 
\begin{equation}
\frac{dI(z)}{dz}=-\rho\sigma_0 I(z),
\end{equation}
from which we recover the simple
\begin{equation}
\sigma_0 n = OD_0 = -\ln I_0/I_f. \label{eq6}
\end{equation}
In the limit that the probe intensity is much stronger than saturation, $I_0\gg I_sat$, even far detuned atoms will absorb light at their maximum, allowing us to again neglect the time dependence and reduce \ref{eq3} to 
\begin{equation}
\frac{dI(z)}{dz}=-\rho\sigma_0 I_{sat}, 
\end{equation}
which integrates to 
\begin{equation}
\sigma_0 n = \frac{I_0 - I_f}{I_{sat}}. \label{eq8}
\end{equation}
We recognize the right hand sides of Eq. (\ref{eq6}) and Eq. (\ref{eq8}) as the two terms in the expression for $OD_1$, Eq. (\ref{eq2}). Thus, in both limits $OD_{corr1}$ should coincide with the value of $OD_0$ obtained by fixing an input atom number $n$ in Eq. (\ref{eq2}). Indeed, they coincide as seen in Fig. \ref{fig:IsatLimits}.
\begin{figure}
	\includegraphics{figure3.pdf}
\caption{Optical density as a function of probe intensity as predicted by the simulation (blue dots) and by Eq. (\ref{eq6}) (green lines), for three different imaging times. The predictions agree in both the high and low intensity limits, and differ for probe intensities comparable to the saturation intensity. The difference is enhanced with increased imaging time.}  
\label{fig:IsatLimits}
\end{figure}
\par We can use the results of this simulation to check if the stationary atom assumption is self consistent, ie if the distance traveled by the atoms in one bin during the imaging time is less than the bin size. However, as can be seen from Fig. \ref{fig:atomTravel}, it turns out that not only do the atoms travel more than the bin size, but they also travel more than the size of the whole cloud, and the back of the cloud even overtakes the front for long enough imaging times. Thus, the atomic distribution as a function of position changes dramatically during the imaging time, and our stationary assumption is completely invalid. 
\begin{figure}
	\includegraphics{figure4.pdf}
\caption{Position of atoms as a function of imaging time for atoms in the first, middle, and last bins of the simulation. The probe intensity here is $1.2 I_{sat}$, and the optical depth is 1.6.}  
\label{fig:atomTravel}
\end{figure}

@@

\STATE $I[n,t]=I[n-1,t] - A I[n-1,t]/(1+S[n].\delta^2+I[n-1,t])$  \COMMENT{Eq. (\ref{eq3})}
\STATE $S[n].\delta\mathrel{+}=B\left(I[n-1,t]-I[n,t]\right)$  \COMMENT{Eq. (\ref{eq4}), detuning in units of $\Gamma/2$} 

@@

\par First, we check the velocity behavior in this model against known limits. One such limit is that of a single supeatom. In this case, there is no attenuation, and the intensity seen by the superatom is constant at $I_0$. The only thing that evolves in time is the detuning of the single superatom. We can differentiate both sides of Eq. (\ref{eq4}) to obtain

@@

and Eq. (\ref{eq3}) simplifies to

@@

Combining Eq. (\ref{eq9}) and Eq. (\ref{oneAtom}), and converting to a dimensionless form, we obtain

@@

where $\Delta = 2\delta/\Gamma$, $\tilde{I} = I_0/I_{sat}$ and $k$ is the photonic wavevector. Equation (\ref{eq10}) can be solved numerically, and is in good agreement with our simulation, as seen in Fig. \ref{fig:oneAtomVel}.

@@

\par We use this model to study the time evolution of the cloud shape during imaging. This can be visualized as the phase space evolution of superatoms, as seen in Fig. \ref{fig:phaseSpace}. We see that the cloud shape is actually strongly distorted during the imaging time. 

@@

\par We compare the optical depths predicted by each of the two models, $OD_{corr1}$ and $OD_{corr2}$. We find that, depite the significant changes in atomic distribution during the imaging time, the predicted optical densities are so slightly changed by including this effect that it is practically undetectable, as seen in Fig. \ref{fig:atomTravel}. In fact, the difference $\left|OD_{corr1}-OD_{corr2}\right|/OD_{corr1} \le 0.005$. Thus, for the purposes of deducing the atom density from experimental optical depths, simply using a stationary model is sufficient. Furthermore, the actual atomic distribution $\rho(z)$ is largely irrelevant, and the only observable is the 2-D atomic density $n=\int\rho(z)\mathrm{d}z$. 

@@

\subsection{Signal to noise optimization}
This simulation can be used to interpret experimentally obtained initial and final intensities. For a given imaging time, we create a look-up table of predicted optical depth as a function of probe intensity and atomic density. We can then find the observed optical depth on this table, with the given probe intensity, and infer the atomic density. The uncertainty in the measured intensities can be propagated through this procedure, and we can establish optimal imaging parameters to maximize the signal to noise ratio of this detection scheme. 
\par Here, the only source of measurement uncertainty we consider is the photon shot noise, which is Poisson distributed and has a standard deviation proportional to $\sqrt{N_p}$, where $N_p$ is the photon number. We then propagate this uncertainty through our correction scheme to get the uncertainty in our deduced value of $\sigma_0 n$. We define the signal to noise ratio (SNR) as $\sigma_0 n/\delta_{\sigma_0 n}$, where $ \delta_{\sigma_0 n}$ is the propagated measurement uncertainty.
\par As seen in Fig. \ref{fig:SNR}, after about 40\us{} extending the imaging time no longer yields appreciable improvement in signal to noise ratio. The advantage of imaging for 40\us{} as opposed to 10\us{} where the uncorrected model is appropriate is about a factor of  1.5 increase in SNR. We performed the experiments described in the second section at 40\us{} imaging time.
\begin{figure}
	\subfigure[]{\includegraphics{figure9a.pdf}}
	\subfigure[]{\includegraphics{figure9b.pdf}}
\caption{Signal to noise ratio (SNR) for three different optical depths after correcting for recoil induced detuning. a. SNR as a function of probe intensity $I_0/I_{sat}=5.0$ and b. SNR as a function of probe intensity for an imaging time of 50\us{}.}  
\label{fig:SNR}
\end{figure}

\subsection{Calibration of saturation intensity}
A key component of good detection is a well calibrated measurement apparatus. In this case, the measurement apparatus is a CCD camera used to take the absorption images. We used Point Grey's Flea3 camera.  Each camera pixel  converts tho photons it is exposed to, with some efficiency, into photoelectrons, and then to a digital signal, and integer we will call the number of 'counts'. The number of photons that hit a pixel and the number of 'counts' it outputs are directly proportional. However, the proportionality constant depends on many factors, such as the quantum efficiency of the camera, the photoelectric conversion factor of the camera, and the polarization of the probe light. 
\par The most accurate way to determine this factor is through a direct experiment. In the limit where the system is adequately described by Eq. (\ref{eq6}), only the ratio of the initial and final intensities matters, and thus this proportionality constant is irrelevant. In all other regimes, however, the ratio of the initial and final intensities to the saturation intensity also comes into play, making the proportionality constant significant. One way to approach this calibration is to determine the value of the saturation intensity in 'counts' per unit time. 
\par In order to calibrate the saturation intensity in camera counts per unit time, we take absorption images of a cloud of $^{40}K$ atoms at three different imaging times, 40us, 100us, and 200us, at varying probe intensities. We pick a small square in the center of the cloud, where the atomic density is approximately uniform, and average the initial and final intensities of each pixel in the square. Thus, for each image we obtain one $I_0$ and one $I_f$, in counts per microsecond. We then do a least squares fit of $OD_{corr}$, our simulated optical depth, to the data. The two fit parameters are the atomic density $n$ at the center of the cloud and the value of $I_{sat}$ in counts per microsecond. As seen in Fig. \ref{fig:isatCalib}, the model produces a good fit to the experimental data, and we obtain a calibration of the saturation intensity for our experiment. 
\begin{figure}
	\includegraphics{figure8.pdf}
\caption{The optical depth as a function of probe intensity for three imaging times: $t=40$\us{} (blue),  $t=75$\us{} (green),  $t=100$\us{} (red). The dots represent experimental data and the lines represent the best fit of simulated data. The optimal fit parameters pictured are a $\sigma_0 n$ of 1.62 and saturation intensity of 29 counts/\us{}. }  
\label{fig:isatCalib}
\end{figure}

@@

\par We first corrected each image for recoil induced detuning using the simulation described above. We simulated a look-up table of predicted optical depths as a function of atomic column density and probe intensity. From our two absorption images we knew the probe intensity and optical depth for each pixel. We then used the simulated look-up table in an inverted fashion to deduce the atomic column density. An example of the effect of this correction procedure is shown in Fig. \ref{fig:SampleCorrection}.  To improve the signal and compensate for our shot to shot number fluctuations, we took 15 equivalent images for each data point and, after correcting, added them together.
